First paper - Draft

### 2017-06-20 paper/paper.Rnw
Continued writing the data description

\usepackage{pdflscape}
\usepackage{subcaption}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[round, colon]{natbib}
\usepackage[colorlinks]{hyperref}
\AtBeginDocument{%
  \hypersetup{
    citecolor=blue,
    linkcolor=blue,   
    urlcolor=blue}}

@@

Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footenote{Even after weighting appropriately.}. Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations. Corak and Waldfogel (2015) don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.

One important drawback of these studies is that they don't present the uncertainty of this estimate. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen & Jonsson 2007). Virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).

@@

One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula. These four countries have no major jump in school selection, quite the opposite to the European average countries. A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. Moreover, Mickleweight & Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.

@@

The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright & Schnepf, 2007)

Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight & Schnepf did it). They also find that France is a lower dispersion country (but in my results this is changing a lot). They also find that Germany and US are very dispersed countries (quite the opposite)

Mickleweight & Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data. Mickleweight & Schnpef find
that average achievement is negatively related to dispersion.

@@

Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
Talk more about how countries with high social mobility has been linked to smallest achievement gaps
Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.

@@

@@

@@

<<variance_pisa>>=

@@

<<>>=

country_scores <-
  map(adapted_year_data, ~ {
  if (unique(.x$wave) == "pisa2015") {
    intsvy::pisa2015.mean.pv("MATH", by = "country", data = .x)
  } else {
    intsvy::pisa.mean.pv("MATH", by = "country", data = .x)
  }
}) %>% setNames(seq(2000, 2015), 3) %>%

>>>>

\usepackage{geometry}
 \geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
 }
\usepackage{pdflscape}
\usepackage{subcaption}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[round, colon]{natbib}
\usepackage[colorlinks]{hyperref}
\AtBeginDocument{%
  \hypersetup{
    citecolor=blue,
    linkcolor=blue,   
    urlcolor=blue}}

@@

Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}. Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations. Corak and Waldfogel (2015) don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.

One important drawback of these studies is that they don't present the uncertainty of this estimate. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen \& Jonsson 2007). Virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).

@@

One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula. These four countries have no major jump in school selection, quite the opposite to the European average countries. A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. Moreover, Mickleweight \& Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.

@@

The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright \& Schnepf, 2007)

Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight \& Schnepf did it). They also find that France is a lower dispersion country (but in my results this is changing a lot). They also find that Germany and US are very dispersed countries (quite the opposite)

Mickleweight \& Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data. Mickleweight \& Schnpef find
that average achievement is negatively related to dispersion.

@@

Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
Talk more about how countries with high social mobility has been linked to smallest achievement gaps
Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.

\section{Research questions}

@@

<<country_sample_numeric_vec>>=
country_rows <- map_dbl(adapted_year_data, nrow)
@

In order the study the above mentioned questions I will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting in the year 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \Sexpr{country_rows[1]}, for PISA 2003, \Sexpr{country_rows[2]}, for PISA 2006, \Sexpr{country_rows[3]}, for PISA 2009, \Sexpr{country_rows[4]}, for PISA 2012, \Sexpr{country_rows[5]}, and for PISA 2015 \Sexpr{country_rows[6]}. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough time to visualize changes in the structure of an educational system.

In order to measure family background PISA includes several variables. Firstly, the ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education (Reardon 2011). This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding. Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time. For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.

Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI). This variable captures the economic status of the family, the closest it can without asking for income information. It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature(Anna study about world-wide inequality). PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

But the most importat variable for our study is the is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (Education at a Glance, OECD, Paris, 2002, Glossary). This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves. In fact, we use the newly-released ESCS index (% http://www.oecd-ilibrary.org/education/pisa-2015-results-volume-i/pisa-2015-technical-background_9789264266490-13-en
) which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.

Talk about plausible values, what they mean and how they're measured.

@@

The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps(Fryer and Levitt 2004). The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, Mickleweight \& Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. Guan Kung Saw (2015), for instance, used parental education as a proxy of SES, whereas Soo-yong Byun and Kyung-keun Kim (2010) use a similar SES index as ours, but created by them.

Reardon (2011) and subsequently (all reardon papers), used a different method developed by Reardon (2011) which we partially adopt adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.

We start by standardizing achievement for each country, year and subject. Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score. We then fit a linear model for each wave where we regress the single achievement score on the age variable. We take the adjusted test scores and divide them by the root mean square error and divide it by the square root of the reliability of the survey for that year.

Due to the standardization of achievement described above, gaps will also be attenuated in years where achievement is less reliably measured. Therefore, gaps are adjusted according to each country’s test reliability for each study, as published in the corresponding technical reports, as well as according to the estimated reliability of each SES report. For studies where both students and parents reported the same SES variable, reliability can be calculated from the microdata. These reliabilities are then applied to all other years.

% To remember the models and formula for the above paragraph.
      % mod1 <- lm(dv ~ AGE,
      %            weights = .x[[weights_var]],
      %            data = .x,
      %            na.action = "na.exclude")
      % 
      % # Take residuals of model and divide by rmse. Multiply that by
      % # 1 / sqrt(reliability of each survey), which is .y in the loop.
      % .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
      % 
      % mod2 <-
      %   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
      %        data = .x,
      %        weights = .x[[weights_var]])

@@

<<variance_pisa, cache = TRUE>>=

@@

<<graphing_achievement_disparity, cache = T>>=

country_scores <-
  map(adapted_year_data, ~ {
  if (unique(.x$wave) == "pisa2015") {
    intsvy::pisa2015.mean.pv("MATH", by = "country", data = .x)
  } else {
    intsvy::pisa.mean.pv("MATH", by = "country", data = .x)
  }
}) %>% setNames(seq(2000, 2015, 3))

### 2017-06-20 paper/paper.tex
Continued writing the data description

\section{Descriptives}

@@

\section{Analysis}

@@

Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footenote{Even after weighting appropriately.}. Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations. Corak and Waldfogel (2015) don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.

One important drawback of these studies is that they don't present the uncertainty of this estimate. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust these numbers.

Say more above Corak and Waldfogel above (book).

Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences and cannot be easily compared. Their findings are very reliable but should be taken as suggestive.  For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries. These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.

Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
Talk more about how countries with high social mobility has been linked to smallest achievement gaps
Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.

\section{Research design}

Make sure PISA variation is constant over time
\end{document}

>>>>

Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}. Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations. Corak and Waldfogel (2015) don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.

One important drawback of these studies is that they don't present the uncertainty of this estimate. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen \& Jonsson 2007). Virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).

One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula. These four countries have no major jump in school selection, quite the opposite to the European average countries. A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. Moreover, Mickleweight \& Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.

Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared. Their findings are very reliable but should be taken as suggestive at best.  For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies. These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.

The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright \& Schnepf, 2007)

Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight \& Schnepf did it). They also find that France is a lower dispersion country (but in my results this is changing a lot). They also find that Germany and US are very dispersed countries (quite the opposite)

Mickleweight \& Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data. Mickleweight \& Schnpef find
that average achievement is negatively related to dispersion.

Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?

Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
Talk more about how countries with high social mobility has been linked to smallest achievement gaps
Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.

\section{Research questions}

\section{Methods}

\subsection{Data}

In order the study the above mentioned questions I will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting in the year 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \ensuremath{1.27388\times 10^{5}}, for PISA 2003, \ensuremath{2.76165\times 10^{5}}, for PISA 2006, \ensuremath{3.9875\times 10^{5}}, for PISA 2009, \ensuremath{5.15958\times 10^{5}}, for PISA 2012, \ensuremath{4.8081\times 10^{5}}, and for PISA 2015 \ensuremath{5.19334\times 10^{5}}. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough time to visualize changes in the structure of an educational system.

In order to measure family background PISA includes several variables. Firstly, the ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education (Reardon 2011). This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding. Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time. For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.

Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI). This variable captures the economic status of the family, the closest it can without asking for income information. It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature(Anna study about world-wide inequality). PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

But the most importat variable for our study is the is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (Education at a Glance, OECD, Paris, 2002, Glossary). This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves. In fact, we use the newly-released ESCS index (% http://www.oecd-ilibrary.org/education/pisa-2015-results-volume-i/pisa-2015-technical-background_9789264266490-13-en
) which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.

Talk about plausible values, what they mean and how they're measured.

\subsection{Data analysis}

The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, Mickleweight & Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education.
\section{Analysis}

\subsection{Descriptives}

@@

\subsection{Results}

@@

\end{document}
